---
title: Problem set 10
date: 2025-04-20
execute: 
  eval: false
---

The data for this problem set is provided by this link: `https://github.com/dmcable/BIOSTAT620/raw/refs/heads/main/data/pset-10-mnist.rds`

Read this object into R. For example, you can use:

```{r}
fn <- tempfile()
download.file("https://github.com/dmcable/BIOSTAT620/raw/refs/heads/main/data/pset-10-mnist.rds", fn)
dat <- readRDS(fn)
file.remove(fn)
```


The object is a list with two components `dat$train` and `dat$test`. Use the data in `dat$train` to  develop a machine learning algorithms to predict the labels for the images in the `dat$test$images` component.

Save the your predicted labels in an object called `digit_predictions`. This should be a vector of integers with length `nrow(dat$test$images)`. It is important that the `digit_predictions` is ordered to match the rows of `dat$test$images`.

Save the object to a file called `digit_predictions.rds` using:

```{r}
library(e1071)

train_images <- dat$train$images
train_labels <- as.factor(dat$train$labels)
train_df <- as.data.frame(train_images)
train_df$label <- train_labels

x_train_full <- train_df[, -ncol(train_df)]
nzv <- apply(x_train_full, 2, sd) != 0
x_train_full <- x_train_full[, nzv]

pca_result <- prcomp(x_train_full, center = TRUE, scale. = TRUE)
train_pca <- as.data.frame(pca_result$x[, 1:200])
train_pca$label <- train_df$label

svm_model <- svm(x = train_pca[, -ncol(train_pca)], y = train_pca$label, kernel = "radial", cost = 5)

test_df <- as.data.frame(dat$test$images)
test_pca <- predict(pca_result, newdata = test_df[, nzv])[, 1:200]
test_pca <- as.data.frame(test_pca)

digit_predictions <- predict(svm_model, newdata = test_pca)

saveRDS(digit_predictions, file = "digit_predictions.rds")
```

You will submit:

1. The file `digit_predictions.rds`

2. A quarto file that reproduces your analysis and provides brief explanations for your choices. 

**If your code reproduces the result**, your grade will be your accuracy  rounded up the closest integer. So, for example, if your accuracy is .993 your grade will be 100%. Depending on the distribution of accuracy values, the teaching staff may issue an update about the grading system used.

You will have one opportunities to redo your predictions after you see your accuracy from your first submission.

